---
title: "02_Feature_Engineering"
author: "Abraham Ybarra"
date: '2023-02-06'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries

```{r cars}
library(tidyverse)
library(here)
library(car)
library(VGAM)
```

## Read in data

```{r}
######## Read in F21 (Ashley) Data ############
# studentSummary_21B <- read.csv(paste0(here(),"/clean_data/F21_Ashley/studentSummary.csv"), header = TRUE)
# late_21B <- read.csv(paste0(here(),"/clean_data/F21_Ashley/lateAssignments.csv"), header = TRUE)
# missing_21B <- read.csv(paste0(here(),"/clean_data/F21_Ashley/missingAssignments.csv"), header = TRUE)
# grades_21B <- read.csv(paste0(here(),"/clean_data/F21_Ashley/gradebook.csv"), header = TRUE)
assignments_21A <- read.csv(paste0(here(),"/clean_data/F21_Ashley/assignments.csv"), header = TRUE, numerals = "no.loss")
```

## Create new variable for time difference between submission time and due date

```{r}
# Read data
# assignments_21A <- read.csv(paste0(here(),"/clean_data/F21_Ashley/assignments.csv"), header = TRUE, numerals = "no.loss")

######## Create time diff variable ############
# Create unique id for each row
assignments_21A <- assignments_21A %>%
  mutate(UID = paste(ANON_ID, ASSIGNMENT_NAME))

# TODO: Figure out the issue with date format for SUBMITTED_AT. Code does not work when SUBMITTED_AT is empty (i.e. no submission)
# even when catching those cases with ifelse(). 
# TEMP FIX: only analyze assignments with submission time, then join back to full dataset
assignments_21A_submitted <- assignments_21A %>%
  filter(SUBMITTED_FLAG == "True")

# Convert date-time values to POSIXct format
assignments_21A_submitted <- assignments_21A_submitted %>%
  mutate(DUE_DATE = as.POSIXct(DUE_DATE, format = "%Y-%m-%d %H:%M:%S", tz = "GMT"),
         SUBMITTED_AT = as.POSIXct(SUBMITTED_AT, format = "%Y-%m-%d %H:%M:%S", tz = "GMT"))

# Compute time difference in hours and store result in new column. Positive means early, negative means late
assignments_21A_submitted <- assignments_21A_submitted %>%
  mutate(HOUR_DIFF = difftime(DUE_DATE, SUBMITTED_AT, units = "hours") %>% as.numeric(),
         .after = SUBMITTED_AT_TIME) %>%
  group_by(ASSIGNMENT_NAME) %>%
  mutate(HOUR_DIFF_NORM = HOUR_DIFF/max(HOUR_DIFF),
         HOUR_DIFF_Z = (HOUR_DIFF - mean(HOUR_DIFF))/sd(HOUR_DIFF),
         LOG_HOUR_DIFF = log(HOUR_DIFF),
         .after = HOUR_DIFF) %>%
  ungroup()
  
# Create yeo-johnson transformation of the HOUR_DIFF variable
assignments_21A_submitted <- assignments_21A_submitted %>%
  mutate(YJ_NEG_2 = yeo.johnson(HOUR_DIFF, lambda = -2),
         YJ_NEG_1 = yeo.johnson(HOUR_DIFF, lambda = -1),
         YJ_0 = yeo.johnson(HOUR_DIFF, lambda = 0),
         YJ_POS_1 = yeo.johnson(HOUR_DIFF, lambda = 1),
         YJ_POS_2 = yeo.johnson(HOUR_DIFF, lambda = 2),)
  

# Add HOUR_DIFF back to the main dataset
assignments_21A <- assignments_21A %>%
  left_join(assignments_21A_submitted %>% 
              select(c(UID, HOUR_DIFF, HOUR_DIFF_NORM, HOUR_DIFF_Z, LOG_HOUR_DIFF, YJ_NEG_2, YJ_NEG_1, YJ_0, YJ_POS_1, YJ_POS_2)), 
            by = "UID") %>%
  relocate(HOUR_DIFF:YJ_POS_2, .after = SUBMITTED_AT_TIME)
```

## Create additional variables

```{r}
######## Create additional variables ############

# Print all assignment names
unique(assignments_21A$ASSIGNMENT_NAME)

# Create variables for assignment type
assignments_21A <- assignments_21A %>%
  mutate(ASSIGNMENT_TYPE = case_when(
    str_detect(ASSIGNMENT_NAME, regex('Short Paper', ignore_case = TRUE)) ~ "Short Paper",
    str_detect(ASSIGNMENT_NAME, regex('Final Exam', ignore_case = TRUE)) ~ "Final Exam",
    str_detect(ASSIGNMENT_NAME, regex('Mid-Term Exam', ignore_case = TRUE)) ~ "Mid-Term Exam",
    str_detect(ASSIGNMENT_NAME, regex('Post-Lecture Quiz', ignore_case = TRUE)) ~ "Post-Lecture Quiz",
    str_detect(ASSIGNMENT_NAME, regex('Pre-Lecture Assignment', ignore_case = TRUE)) ~ "Pre-Lecture Assignment",
    str_detect(ASSIGNMENT_NAME, regex('Study habits challenge', ignore_case = TRUE)) ~ "Study Habits Challenge",
    str_detect(ASSIGNMENT_NAME, regex('Reading Quiz', ignore_case = TRUE)) ~ "Reading Quiz",
    .default = NA),
    .after = ASSIGNMENT_NAME)

# Create variable for assignment score out of 100
assignments_21A <- assignments_21A %>%
  mutate(SCORE_100 = SCORE / POINTS_POSSIBLE * 100, .after = POINTS_POSSIBLE)

# Create PI = {1 if after 75% of students; 0 otherwise} 
assignments_21A <- assignments_21A %>% 
  group_by(ASSIGNMENT_NAME) %>%
  mutate(CUTOFF_75 = format(quantile(as.POSIXct(SUBMITTED_AT, format = "%Y-%m-%d %H:%M:%S"), 0.75, na.rm = TRUE), "%Y-%m-%d %H:%M:%S"),
         PI = as.integer(as.POSIXct(SUBMITTED_AT, format = "%Y-%m-%d %H:%M:%S") >= as.POSIXct(CUTOFF_75, format = "%Y-%m-%d %H:%M:%S")),
         .after = SUBMITTED_AT_TIME) %>%
  ungroup() %>%
  replace_na(list(PI = 1))
```

## Create student df w/ variables for % missing, late, on time, etc, as well as overall scores for each assignment type

```{r}
######## Create variable for % of all assignments submitted late, missing, etc. ############

# Obtain overall score for each assignment type for each student, as well as missing/late stats
studentSummary_21A <- assignments_21A %>%
  group_by(ANON_ID) %>%
  summarise(SECTION = unique(Section.Id),
            OVERALL_COURSE_GRADE = unique(OVERALL_COURSE_GRADE),
            
            SHORT_PAPER_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Short Paper", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Short Paper", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            FINAL_EXAM_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Final Exam", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Final Exam", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            MIDTERM_EXAM_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Mid-Term Exam", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Mid-Term Exam", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            POST_LECTURE_QUIZ_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Post-Lecture Quiz", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Post-Lecture Quiz", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            PRE_LECTURE_ASSIGNMENT_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Pre-Lecture Assignment", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Pre-Lecture Assignment", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            STUDY_HABITS_CHALLENGE_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Study Habits Challenge", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Final Exam", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            READING_QUIZ_SCORE = sum(if_else(ASSIGNMENT_TYPE == "Reading Quiz", SCORE, NA), na.rm = TRUE) / sum(if_else(ASSIGNMENT_TYPE == "Reading Quiz", POINTS_POSSIBLE, NA), na.rm = TRUE) * 100,
            
            NUM_ASSIGNMENTS = n(),
            NUM_MISSING = sum(SUBMITTED_FLAG == "False"),
            NUM_LATE = sum(HOUR_DIFF < 0, na.rm = TRUE),
            NUM_NOT_ON_TIME = NUM_MISSING + NUM_LATE,
            NUM_ON_TIME = NUM_ASSIGNMENTS - NUM_NOT_ON_TIME,
            PCT_MISSING = NUM_MISSING / NUM_ASSIGNMENTS,
            PCT_LATE = NUM_LATE / NUM_ASSIGNMENTS,
            LOG_PCT_LATE = log(PCT_LATE + .01),
            PCT_NOT_ON_TIME = NUM_NOT_ON_TIME / NUM_ASSIGNMENTS,
            LOG_PCT_NOT_ON_TIME = log(PCT_NOT_ON_TIME + .01),
            PCT_ON_TIME = NUM_ON_TIME / NUM_ASSIGNMENTS,
            
            MEAN_HOUR_DIFF = mean(HOUR_DIFF, na.rm = TRUE),
            MEAN_HOUR_DIFF_Z = mean(HOUR_DIFF_Z, na.rm = TRUE),
            
            MEAN_PI = mean(PI, na.rm = TRUE),
            LOG_MEAN_PI = ifelse(is.na(MEAN_PI), NA, log(MEAN_PI + .01)))
```

# Join select summary stats back into assignment dataset

```{r}
assignments_21A <- assignments_21A %>%
  inner_join(studentSummary_21A %>% select(ANON_ID, FINAL_EXAM_SCORE), by = "ANON_ID") %>%
  relocate(FINAL_EXAM_SCORE, .after = OVERALL_COURSE_GRADE)
```


# Validate with data pulled directly from canvas

```{r}
# Calculate missing, late, on time, etc. using data pulled directly from canvas
# missingSummary_21B <- missing_21B %>%
#   group_by(Random.Id) %>%
#   summarise(numMissing = n())
# 
# lateSummary_21B <- late_21B %>%
#   group_by(Random.Id) %>%
#   summarise(numLate = n())
# 
# studentSummary_21B <- studentSummary_21B %>%
#   left_join(missingSummary_21B, by = "Random.Id") %>%
#   left_join(lateSummary_21B, by = "Random.Id") %>%
#   mutate(numAssignmentsTot = 150, # TODO: how many assignments total?
#          numMissing = ifelse(is.na(numMissing), 0, numMissing),
#          pctMissing = numMissing / numAssignmentsTot,
#          numLate = ifelse(is.na(numLate), 0, numLate),
#          pctLate = numLate / numAssignmentsTot,
#          numNotOnTime = numMissing + numLate,
#          pctNotOnTime = numNotOnTime / numAssignmentsTot,
#          numOnTime = numAssignmentsTot - numNotOnTime,
#          pctOnTime = numOnTime / numAssignmentsTot) %>%
#   relocate(c(numAssignmentsTot, numMissing, pctMissing), .before = numLate)
```

## Write to csv

```{r}
######## Write cleaned F21 (Ashley) data to csv file ################
write.csv(studentSummary_21A,paste0(here(),"/final_data/F21_Ashley/studentSummary.csv"), row.names = FALSE)
write.csv(assignments_21A,paste0(here(),"/final_data/F21_Ashley/assignments.csv"), row.names = FALSE)
```





